Paper

### 2025-02-27 paper/index.qmd
reorganise scripts

```{r, include=FALSE}

run_sims <- FALSE

@@

# to change body font size:
# change this line in typst-template.typ:
# set text(font: "STIX Two Text", size: 24pt)
library(tidyverse)
library(abc)
library(lhs)
library(flextable)
source(file.path("..", "calibrate.R"))
# load(file.path("..", "out", "calibrate.RData"))
```

Transmission of potentially pathogenic bacteria under antimicrobial selective pressure has important implications for control of hospital-acquired infection. Examples include *Clostridioides difficile* infection (CDI) and Vancomycin-resistant *Enterococcus*. In addition to patient factors, colonisation risk depends on environmental factors (e.g., surface cleaning), proximity to other colonised patients, and use of antimicrobial agents. To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by Salgado et al @salgado2009.

@@

A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@anderson1991; @keeling2011]:

@@

Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgado2009]. Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchell2020].

@@

Rejection sampling approximate Bayesian computation (ABC) was employed to calibrate the model to the outbreak data [@sunnåker2013]. Priors were determined from domain knowledge or extracted from existing literature. A summary of the priors used is available in @tbl-priors.

@@

| Name                                                       | Distribution | Parameters                                                                                                   | Description                                                                     | Reference                                          |
|---------------|---------------|---------------|---------------|---------------|
| Baseline (pre-outbreak) antimicrobial prescription rate    | Constant     | 49.9%                                                                                                        | Proportion of patients administered antimicrobials during admission             | @versporten2018                                    |
| Admission duration                                         | Poisson      | $\lambda = 4.6  \text{ days}$                                                                                |                                                                                 | @levitkryankelixhauserastrangesekassedccoffeyr2007 |
| Proportion of toilets that are shared between \>1 patients | Constant     | 60%                                                                                                          | The remaining proportion (40%) of toilets were dedicated to one patient         | N/A                                                |
| Community colonisation rate                                | Uniform      | $a = 0.01, b = 0.1$                                                                                          |                                                                                 | @crobach2018                                       |
| Antimicrobial effect                                       | Uniform      | $ a = 1.31, b = 1.87 $                                                                                     | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimings2014                                      |
| CDI incidence                                              | Uniform      | $a=2.8, b=12.1$                               | Per 10,000 patient-days                                                         | @marra2020                                         |
| Mean toilet frequenting rate                               | Uniform      | $a=0.5,b=5$                               | Per day; used as \lambda parameter for Poisson distribution                     |                                                    |
| Mean toilet cleaning rate                                  | Uniform      | $a=0.5, b=3$                               | Per day; used as \lambda parameter for Poisson distribution                     |                                                    |
| Toilet contamination effect                                | Uniform      | $a=0.01,b=0.99$                               | Contamination impact by infected/colonised patients through toilet use          |                                                    |
| Toilet cleaning effect                                     | Uniform      | $a=0.01, b=0.99$                               | Impact of toilet cleaning on contamination                                      |                                                    |

@@

The agent-based model was implemented in NetLogo v6.2.2 [@wilenskyu.1999]. The R package `nlrx` was used to execute simulations and transfer output data for analysis in R [@salecker2019; @rcoreteam2022].

@@

```{r, include=T}
params_table <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble() %>% 
  dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
  dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
  summarise(across(everything(), ~ paste0(
    round(mean(.x), 2),
    " (",
    round(sd(.x), 2),
    ")"
    )))

@@

params_table %>% 
  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
               names_pattern = "(b|o|c)-(.*)") %>% 
  pivot_wider(names_from = period, values_from = value) %>% 
  # reorder cols
  dplyr::select(param, b, o, c) %>% 
  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
  mutate(units = param) %>% 
  relocate(units, .after = param) %>%
  flextable() %>% 
  labelizor(j = "units", part = "body",
            labels = param_units) %>% 
  labelizor(j = "param", part = "body",
            labels = param_names)

params_table %>% 
  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
               names_pattern = "(b|o|c)-(.*)") %>% 
  pivot_wider(names_from = period, values_from = value) %>% 
  # reorder cols
  dplyr::select(param, b, o, c) %>% 
  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
  mutate(units = param) %>% 
  relocate(units, .after = param) %>%
  knitr::kable()
```

@@

# Plot the results
library(ggplot2)
output_long <- reshape2::melt(output_df, id = "time")
ggplot(output_long, aes(x = time, y = value, color = variable)) +
  geom_line(size = 1) +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment") +
  scale_color_manual(values = c("blue", "red"),
                     labels = c("Non-Infected (N)", "Infected (I)")) +
  theme_minimal()

>>>>

```{r setup, include=FALSE}
# here::i_am("paper/index.qmd")
run_sims <- FALSE

@@

# to change body font size:
# change this line in typst-template.typ:
# set text(font: "STIX Two Text", size: 24pt)
library(tidyverse)
library(abc)
library(lhs)
library(flextable)
```

```{r}
# print(here::here())
# file.exists(here::here("scripts", "calibrate.R"))
```

```{r load-data, include=FALSE}
# print(file.exists(here::here("scripts", "calibrate.R")))
# source(here::here("scripts", "calibrate.R"))
source(file.path("..", "scripts", "calibrate.R"), chdir = TRUE)
# load(file.path("..", "out", "calibrate.RData"))
```

Transmission of potentially pathogenic bacteria under antimicrobial selective pressure has important implications for control of hospital-acquired infection. Examples include *Clostridioides difficile* infection (CDI) and Vancomycin-resistant *Enterococcus*. In addition to patient factors, colonisation risk depends on environmental factors (e.g., surface cleaning), proximity to other colonised patients, and use of antimicrobial agents. To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by @salgadoAnalysisOutbreakClostridium2009.

@@

A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:

@@

Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgado2009]. Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchellEngaugeDigitizer2020].

@@

Rejection sampling approximate Bayesian computation (ABC) was employed to calibrate the model to the outbreak data [@sunnakerApproximateBayesianComputation2013]. Priors were determined from domain knowledge or extracted from existing literature. A summary of the priors used is available in @tbl-priors.

@@

| Name | Distribution | Parameters | Description | Reference |
|---------------|---------------|---------------|---------------|---------------|
| Baseline (pre-outbreak) antimicrobial prescription rate | Constant | 49.9% | Proportion of patients administered antimicrobials during admission | @versportenAntimicrobialConsumptionResistance2018 |
| Admission duration | Poisson | $\lambda = 4.6  \text{ days}$ |  | @levitkryankelixhauserastrangesekassedccoffeyrHCUPFactsFigures2007 |
| Proportion of toilets that are shared between \>1 patients | Constant | 60% | The remaining proportion (40%) of toilets were dedicated to one patient | N/A |
| Community colonisation rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
| Antimicrobial effect | Uniform | \$ a = 1.31, b = 1.87 \$ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
| CDI incidence | Uniform | $a=2.8, b=12.1$ | Per 10,000 patient-days | @marraIncidenceOutcomesAssociated2020 |
| Mean toilet frequenting rate | Uniform | $a=0.5,b=5$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Mean toilet cleaning rate | Uniform | $a=0.5, b=3$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Toilet contamination effect | Uniform | $a=0.01,b=0.99$ | Contamination impact by infected/colonised patients through toilet use |  |
| Toilet cleaning effect | Uniform | $a=0.01, b=0.99$ | Impact of toilet cleaning on contamination |  |

@@

The agent-based model was implemented in NetLogo v6.2.2 [@wilenskyu.NetLogo1999]. The R package `nlrx` was used to execute simulations and transfer output data for analysis in R [@rcoreteamLanguageEnvironmentStatistical2022; @saleckerNlrxPackageNextgeneration2019].

@@

```{r, include=T}
#| tbl-cap: "Summary of parameter posteriors after calibration using Approximate bayesian computation."
#| label: tbl-posteriors
#| echo: false
params_table <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble() %>% 
  dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
  dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
  summarise(across(everything(), ~ paste0(
    round(mean(.x), 2),
    " (",
    round(sd(.x), 2),
    ")"
    )))

@@

params_table %>% 
  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
               names_pattern = "(b|o|c)-(.*)") %>% 
  pivot_wider(names_from = period, values_from = value) %>% 
  # reorder cols
  dplyr::select(param, b, o, c) %>% 
  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
  mutate(units = param) %>% 
  relocate(units, .after = param) %>%
  flextable() %>% 
  labelizor(j = "units", part = "body",
            labels = param_units) %>% 
  labelizor(j = "param", part = "body",
            labels = param_names) %>% 
  width(width = 1) %>% 
  font(fontname="Times New Roman",
       part = "all") %>% 
  fontsize(size = 10, part = "all")
```

@@

# Plot the results
library(ggplot2)
output_long <- reshape2::melt(output_df, id = "time")
